---
title: "607_Project2A_DylanGold"
format: html
editor: visual
---

## Codebase #1

### Importing data

In this file I will tidy and analyze the dataset I posted myself in discussion 5A. This source is at https://www.kaggle.com/datasets/eugeniyosetrov/renewable-power-plants?resource=download&select=renewable_capacity_timeseries.csv\

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r,message=FALSE, warning=FALSE}
url <- "https://raw.githubusercontent.com/DylanGoldJ/607-Project-2/refs/heads/main/FileA/renewable_capacity_timeseries.csv"

df <- read_csv(
  file = url,
  col_names = TRUE
)
# Head is 0 values, show tail for non 0 values.
tail(df, 8)
```

This data set has the capacity of different types of energy sources in different countries.

### Tidying the data

I will use glimpse to see the formatting of the columns.

```{r}
glimpse(df)
```

We can see that there are 40 columns when we realistically just need 4 columns. We can try to convert our columns into day, country, energy_type, energy_amount. I will try to separate the names by the first appearance of '\_'.

```{r}
energy_data <- df %>%
  pivot_longer(
    cols = !c("day"),
    names_to = c("country", "energy_type"),
    names_pattern = "^([^_]+)_(.*)", # This regex is starting with : first grouping - characters that are not _, then _, second grouping - rest of string.
    values_to = "energy_amount"
  )

tail(energy_data, 8)

```

We now have our data in a longer format with just 4 columns. I will check to see NA values that we may need to supplement.

```{r}
colSums(is.na(energy_data))
```

We don't have an NA values. Now that the data is in a tidy format we can start to analyze it.

### Analysis

My first interest is finding the
